---
title: "Penguins Dataset Analysis"
author: "Sourena Mohit"
date: "`r Sys.Date()`"
output: html_notebook
---

# **Introduction**
The `palmerpenguins` dataset provides information about three species of penguins from the Palmer Archipelago.  
In this analysis, we will:
- Load the dataset  
- Inspect the data structure  
- Perform basic exploratory data analysis (EDA)  

---

## **1. Load the Dataset**
```{r setup, message=FALSE, warning=FALSE}
# install.packages("palmerpenguins")  # Install package (only needed once)
library(palmerpenguins)  # Load the package
data("penguins")  # Load the dataset
```

---

## **2. Data Inspection & Summary**
```{r data-inspection}
# View first 10 rows
head(penguins, 10)

# Summary statistics
summary(penguins)

# List all variables (column names)
names(penguins)

# Count the number of observations (rows)
nrow(penguins)

# Count the number of variables (columns)
ncol(penguins)

# Get dataset dimensions (rows & columns)
dim(penguins)

# Check data types of each column
str(penguins)

# Count unique values in categorical columns
table(penguins$species)
table(penguins$sex)
table(penguins$island)
```

---

## **3. Handling Missing Values & Data Cleaning**
```{r missing-values}
# Identify and count rows with missing values
sum(!complete.cases(penguins))  # Total missing rows

# Check missing values per column
colSums(is.na(penguins))

# Remove rows with missing values
clean_penguins <- na.omit(penguins)

# Verify changes
nrow(penguins)       # Original dataset size
nrow(clean_penguins)  # Size after removing missing values
```

---

## **4. Basic Statistics & Data Manipulation**
```{r statistics}
# Calculate mean and median of flipper length column (handling missing values)
mean(penguins$flipper_length_mm, na.rm = TRUE)
median(penguins$flipper_length_mm, na.rm = TRUE)

# Filter penguins of species "Adelie" with flipper length > 200mm
filtered_penguins <- subset(penguins, species == "Adelie" & flipper_length_mm > 200)

# Create a new BMI column (Body Mass Index based on flipper length)
penguins$BMI <- penguins$body_mass_g / (penguins$flipper_length_mm ^ 2)
clean_penguins$BMI <- clean_penguins$body_mass_g / (clean_penguins$flipper_length_mm ^ 2)

# Display first 10 BMI values
head(penguins$BMI, 10)
head(clean_penguins$BMI, 10)
```
**Explanation:**
- `mean()` calculates the average value of a numeric column, and `na.rm = TRUE` ensures missing values are ignored.
- `median()` finds the middle value, which is useful for skewed distributions.
- `subset()` is used to filter rows based on conditions (species is "Adelie" and flipper length > 200 mm).
- `BMI` (Body Mass Index) is calculated as `body_mass_g / flipper_length_mm^2`, creating a new column.

---

## **5. Data Visualization**
```{r visualization}
# Histogram of Body Mass
hist(penguins$body_mass_g, main = "Penguin Body Mass Distribution", xlab = "Body Mass (g)", col = "lightblue")
hist(clean_penguins$body_mass_g, main = "Cleaned Penguin Body Mass Distribution", xlab = "Body Mass (g)", col = "lightgreen")

# Boxplot of Flipper Length
boxplot(penguins$flipper_length_mm, main = "Flipper Length Distribution", ylab = "Flipper Length (mm)", col = "orange")

# Scatter Plot: Flipper Length vs. Body Mass
plot(
  penguins$flipper_length_mm, penguins$body_mass_g,
  main = "Flipper Length vs Body Mass",
  xlab = "Flipper Length (mm)", ylab = "Body Mass (g)",
  col = "blue", pch = 16
)

# Bar Plot: Count of Penguins by Species
barplot(table(penguins$species), main = "Count of Penguins by Species", ylab = "Count", col = c("lightblue", "lightgreen", "pink"))
```
---

## **6. Data Grouping and Summarization**
```{r grouping}
# Aggregate function: Compute mean flipper length per species
aggregate(flipper_length_mm ~ species, data = penguins, mean , na.rm = TRUE)

# Count number of penguins per species-sex combination
table(penguins$species , penguins$sex)

# Summarize multiple variables (flipper length and body mass) per species
aggregate(cbind(flipper_length_mm, body_mass_g) ~ species, data = penguins , 
          function(x) c(mean = mean(x, na.rm = TRUE) , sd = sd(x , na.rm = TRUE)))
```
**Explanation:**
- `aggregate()` allows performing grouped operations. Here, we compute mean values per species.
- `table()` is used to count occurrences of categorical values (species-sex combination).
- `cbind()` combines multiple variables to allow summary operations on both at the same time.

---
## **7. Data Analysis**
```{r DA}
# Variance of flipper length
var(penguins$flipper_length_mm, na.rm = TRUE)
var(clean_penguins$body_mass_g)

# Covariance between flipper length and body mass
cov(penguins$flipper_length_mm, penguins$body_mass_g, use = "complete.obs")

# Correlation between flipper length and body mass
cor(penguins$flipper_length_mm, penguins$body_mass_g, use = "complete.obs")
```
**Explanation:**
- `var()` calculates variance, which measures the spread of data.
- `cov()` measures the direction of linear relationships between two variables.
- `cor()` calculates correlation, which standardizes covariance and ranges from -1 to 1, showing the strength and direction of a relationship.
